Bird beak's elaboration and innovation

### 2021-12-02 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
## Plotting just the space
covar.plot(superorder_shapespace,
           main = "Shapespace",
           cex = 0.5,
           legend = TRUE,
           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
           legend.args = list(x = "bottomright", cex = 2/3))
```

@@

```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
par(mfrow = c(2,2))
## Plotting mean axes
covar.plot(superorder_shapespace,
           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average ellipses by super order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5)
## Plotting mean ellipses
covar.plot(superorder_shapespace,
           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average major axes by super order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5)
covar.plot(order_shapespace,
           col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
           main = "Average ellipses by order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5)
## Plotting mean ellipses
covar.plot(order_shapespace,
           col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
           main = "Average major axes by order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5)
```

@@

    ## Which will be the centre plot?
    centre <- ceiling(n/2)
    ## Plot everything
    op <- par(mfrow = c(1, n+1))
    ## Plot the first panel
    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
    empty <- matrix(0, ncol = length(results[[1]]$disparity))
    colnames(empty) <- names(results[[1]]$disparity)[rank]
    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
    for(i in 1:n) {
        ## Is position?
        is_position <- names(results)[[i]] == "position"
        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
    }
    par(op)
}
wrap.plot(results$super_results_between,
              col = rainbow(n.subsets(superorder_shapespace)-1),
              main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
              names = c("elaboration", "innovation"))
```

@@

```{r, echo = FALSE, fig.height = 8, fig.width = 8}
wrap.plot(results$super_results_between,
          xlim = c(0,6),
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```
</details>

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
wrap.plot(results$order_results_between,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, echo = FALSE, fig.height = 16, fig.width = 8}
wrap.plot(results$order_results_between,
          xlim = c(0, 15),
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```
</details>

@@

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot the results
wrap.plot(results$super_results_phylo,
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$order_results_phylo,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 8, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$super_results_within,
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "[species on group] superorders species\nonto their group's VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$order_results_within,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "[species on group] orders species\nonto their group's VCV",
          names = c("elaboration", "innovation"))
```

@@

## Plotting one correlation 
plot.correlations <- function(cor.results, col.sub, col.fun = rainbow, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
    dots <- list(...)

@@

@@

@@

>>>>

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
## The ggplot color panel
gg.color.hue <- function(n) {
    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
}

# colour.palette <- rainbow
colour.palette <- gg.color.hue

## Plotting just the space
covar.plot(superorder_shapespace,
           main = "Shapespace",
           cex = 0.5,
           legend = TRUE,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           legend.args = list(x = "bottomright", cex = 2/3))
```

@@

```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
par(mfrow = c(2,2))
## Plotting mean axes
covar.plot(superorder_shapespace,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average ellipses by super order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
## Plotting mean ellipses
covar.plot(superorder_shapespace,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average major axes by super order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
covar.plot(order_shapespace,
           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
           main = "Average ellipses by order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
## Plotting mean ellipses
covar.plot(order_shapespace,
           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
           main = "Average major axes by order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
```

@@

    ## Which will be the centre plot?
    centre <- ceiling(n/2)
    ## Plot everything
    op <- par(mfrow = c(1, n+1))
    ## Plot the first panel
    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
    empty <- matrix(0, ncol = length(results[[1]]$disparity))
    colnames(empty) <- names(results[[1]]$disparity)[rank]
    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
    for(i in 1:n) {
        ## Is position?
        is_position <- names(results)[[i]] == "position"
        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
    }
    par(op)
}
wrap.plot(results$super_results_between,
              col = colour.palette(n.subsets(superorder_shapespace)-1),
              main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
              names = c("elaboration", "innovation"))
```

@@

```{r, echo = FALSE, fig.height = 8, fig.width = 8}
wrap.plot(results$super_results_between,
          xlim = c(0,6),
          col = colour.palette(n.subsets(superorder_shapespace)-1),
          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```
</details>

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
wrap.plot(results$order_results_between,
          col = colour.palette(n.subsets(order_shapespace)-1),
          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, echo = FALSE, fig.height = 16, fig.width = 8}
wrap.plot(results$order_results_between,
          xlim = c(0, 15),
          col = colour.palette(n.subsets(order_shapespace)-1),
          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```
</details>

@@

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot the results
wrap.plot(results$super_results_phylo,
          col = colour.palette(n.subsets(superorder_shapespace)-1),
          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$order_results_phylo,
          col = colour.palette(n.subsets(order_shapespace)-1),
          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 8, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$super_results_within,
          col = colour.palette(n.subsets(superorder_shapespace)-1),
          main = "[species on group] superorders species\nonto their group's VCV",
          names = c("elaboration", "innovation"))
```

@@

```{r, fig.height = 16, fig.width = 8, echo = FALSE}
## Plot the results
wrap.plot(results$order_results_within,
          col = colour.palette(n.subsets(order_shapespace)-1),
          main = "[species on group] orders species\nonto their group's VCV",
          names = c("elaboration", "innovation"))
```

@@

## Plotting one correlation 
plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
    dots <- list(...)

@@

<details>
  <summary>[Click to expand] Finding corner cases groups.
</summary>

So here's how you remove exploration and elaboration: by centering these three species on the centre of the phylo axis:

Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
colour.palette <- rainbow
posteriors <- shapespace_allbirds_lvl_superorder_order

# plot(superorder_shapespace$matrix[[1]][, 1:2], pch = 19, cex = 0.5)
# identify(superorder_shapespace$matrix[[1]][, 1:2])
# shapespace[c(268, 286, 80), c(9, 10)]

## Moving these three species around: Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
fixed_points <- c(268, 286, 408)
skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
move_around <- c(8749, 8750, 8751)

## Get the major axis
phylo_axis <- axis.covar(superorder_shapespace, sample = median)$phylogeny[[1]]
## Point with no elaboration/exploration
no_nothing <- 0

## Example with no exploration/elaboration
skewed_space[move_around, c(1:8)] <- no_nothing

## Recreate the space
superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
```

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Visualise all that
par(mfrow = c(1,2))
## Plot the space
covar.plot(superorder_skewed_space,
           cex = 0.5,
           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
           main = "elements without\nelaboration innovation")
## Add the phylogenetic axis
lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
## Adding the points
points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)

## Recalculate the disparity
results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
## Getting the correlation plot
col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
cor_results_skewed <- get.correlations(results_skewed)
plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
```

So easy right? We just need to change them along the PC axis to move them around in space.

No, it's not fucking easy because it's a 8 dimensional problem... Even though the two first axes cumulate 86.27% of the variance, fixing the three elements on the two elements and centering them on the 6 other dimensions (13.73% of the variance) results in BIG changes:

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Reinitialising the space
skewed_space <- rbind(shapespace, shapespace[fixed_points, ])

## Example with no exploration/elaboration
skewed_space[move_around, c(3:8)] <- 0

## Recreate the space
superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
```

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Visualise all that
par(mfrow = c(1,2))
## Plot the space
covar.plot(superorder_skewed_space,
           cex = 0.5,
           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
           main = "Centring only on dimensions 3:8")
## Add the phylogenetic axis
lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
## Adding the points
points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)

## Recalculate the disparity
results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
## Getting the correlation plot
col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
cor_results_skewed <- get.correlations(results_skewed)
plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
```

For example, the blue element, Anas_falcata is moved to no elaboration/exploration at all if we do that ignoring data past the second axis.
There's a duck axis lurking in the depth... (spoiler, it's the fourth dimension).

Let's flatten the problem in 2D for simplification:

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Reinitialising the space
skewed_space <- rbind(shapespace, shapespace[fixed_points, ])

## Recreate the space
superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #

## 2Ding the space
dimensions <- c(1,2)
superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
superorder_skewed_space$call$dimensions <- dimensions
fun2d <- function(covar) {
    covar$VCV <- covar$VCV[dimensions, dimensions]
    covar$loc <- covar$loc[dimensions]
    return(covar)
}
superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
```

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Visualise all that
par(mfrow = c(1,2))
## Plot the space
covar.plot(superorder_skewed_space,
           cex = 0.5,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Flattening the results")
## Add the phylogenetic axis
lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
## Adding the points
points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)

## Recalculate the disparity
results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
## Getting the correlation plot
col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
cor_results_skewed <- get.correlations(results_skewed)
plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
abline(a = 0, b = 30, col = "grey")
abline(a = 0, b = 0.03, col = "grey")
```

So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.03 and 30).
Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Reinitialising the space
skewed_space <- rbind(shapespace, shapespace[fixed_points, ])

## Move around the first species to have elaboration = 0 and innovation = 0
skewed_space[move_around[1], c(1:2)] <- 0
## Move around the second species to have elaboration = 0 and innovation = 1
skewed_space[move_around[2], c(1:2)] <- c(0, 0.4)
## Move around the third species to have elaboration = 1 and innovation = 0
skewed_space[move_around[3], c(1:2)] <- c(0.4, 0)

## Recreate the space
superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #

## 2Ding the space
dimensions <- c(1,2)
superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
superorder_skewed_space$call$dimensions <- dimensions
fun2d <- function(covar) {
    covar$VCV <- covar$VCV[dimensions, dimensions]
    covar$loc <- covar$loc[dimensions]
    return(covar)
}
superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
```

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
## Visualise all that
par(mfrow = c(1,2))
## Plot the space
covar.plot(superorder_skewed_space,
           cex = 0.5,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Flattening the results")
## Add the phylogenetic axis
lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
## Adding the points
points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)

## Recalculate the disparity
results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
## Getting the correlation plot
col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
cor_results_skewed <- get.correlations(results_skewed)
plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
abline(a = 0, b = 30, col = "grey")
abline(a = 0, b = 0.03, col = "grey")
```

TODO: fix this!

@@

Regardless, here are the results when projecting the species on their own group's VCV

@@

<!-- THIS BIT IS IGNORED FOR NOW -->
